---
title: "Quick start"
author: "The Health Foundation Analytics Lab"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



## Discover what is available:

There are a few helper functions to help you find out what data sets are available as well as the corresponding editions and versions The `ons_available_datasets()` function will return a dataframe with information about all available datasets. The `id` column is what you need to download a dataset.  
<!-- Important that eval = FALSE here, otherwise building the vignette downloads data from the ONS API, which is not desirable. -->
```{r , eval = FALSE, include = TRUE}

datasets <-  ons_available_datasets()

 datasets() %>% 
  select(id)
                                    id
1                               cpih01
2                     mid-year-pop-est
3                   ashe-table-7-hours
4                ashe-table-7-earnings
5                   ashe-table-8-hours
6                ashe-table-8-earnings
7                           opss-rates
8                      opss-membership
9                wellbeing-year-ending
10           wellbeing-local-authority
...


```

Once you have picked a dataset, you need pick the edition you want. This can be done using `ons_available_editions()`.
```{r, eval = FALSE, include = TRUE}

# Discover the available editions for a particular dataset
ons_available_editions(id = "mid-year-pop-est")

edition
<chr>
1	mid-2018-april-2019-geography			
2	mid-2019-april-2020-geography			
3	time-series



```

Finally, you need to find out what what versions are availble for a specific edition of a dataset. 
```{r, eval = FALSE, include = TRUE}
# Discover the available versions for a particular edition

ons_available_versions("mid-year-pop-est", "time-series")

  version
1       1
2       2
3       3
4       4

```

## Download the data

You should now be ready to download the data. Start by specifying where you want the data to downloaded to. The `monstr_pipeline_defaults()` returns a default folder structure (without creating it). You can specify the a file path base using the `download_root` argument. If you do not specify `download_root`, the base file path will be your project root if you are using Rstudio projects and wherever you working directory is set to otherwise. The output from `monstr_pipeline_defaults()` is then fed to `ons_datasets_setup()` which queries the ONS API to get the relevant information to prepare for downloading the data. Finally, `ons_download()` downloads the data. The rest of the piped code reads in, cleans and saves a clean version of the data. 
```{r, eval=FALSE, include=TRUE}
monstr_pipeline_defaults(download_root="/path/to/download/root/") %>% 
  ons_datasets_setup() %>% # Uses the monstr 'standards' for location and format
	ons_dataset_by_id("weekly-deaths-local-authority") %>%
  ons_download(format="csv") %>%
	monstr_read_file() %>%  
	monstr_clean() %>%
	monstr_write_clean(format="all")
  
```




## Quickstart

<!-- Important that eval = FALSE here, otherwise building the vignette downloads data from the ONS API, which is not desirable. -->
```{r , eval = FALSE, include = TRUE}

library(monstR)
library(magrittr)

  # ons_* fns relate only to ONS datasource
  # monstr_* fns are general purpose for all datasources
> ons_datasets_setup(monstr_pipeline_defaults()) %>% # Uses the monstr 'standards' for location and format
	ons_dataset_by_id("weekly-deaths-local-authority") %>%
	ons_download(format="csv") %>%
	monstr_read_file() %>%  
	monstr_clean() %>%
	monstr_write_clean(format="all")

   # latest version of 'weekly-deaths-local-authority' will be available as a csv 
   # below {{root}}/data/raw/...
   
   # Furthermore, a 'cleaned' version of the dataset will be available as a csv,xlsx, and rds
   # below {{root}}/data/clean/...
```


## Discover what is available:

<!-- Important that eval = FALSE here, otherwise building the vignette downloads data from the ONS API, which is not desirable. -->
```{r , eval = FALSE, include = TRUE}
> ons_available_datasets()
                                    id
1                               cpih01
2                     mid-year-pop-est
3                   ashe-table-7-hours
4                ashe-table-7-earnings
5                   ashe-table-8-hours
6                ashe-table-8-earnings
7                           opss-rates
8                      opss-membership
9                wellbeing-year-ending
10           wellbeing-local-authority
...


# Discover the available editions for a particular dataset
> ons_available_editions("regional-gdp-by-quarter")

      edition
1 time-series


# Discover the available versions for a particular edition
> ons_available_versions("regional-gdp-by-quarter", "time-series")

  version
1       1
2       2
3       3
4       4
5       5
```

## Further Examples

### Download the latest weekly-deaths-local-authority data as a csv.

<!-- Important that eval = FALSE here, otherwise building the vignette downloads data from the ONS API, which is not desirable. -->
```{r , eval = FALSE, include = TRUE}
ons_datasets_setup(monstr_pipeline_defaults()) %>%
	ons_dataset_by_id("weekly-deaths-local-authority") %>%
	ons_download(format="csv")

# file will be in `{{root}}/data/raw/ons/weekly-deaths-local-authority/time-series/vN.csv`
# metadata about the file will be in `{{root}}/data/raw/ons/weekly-deaths-local-authority/time-series/vN.csv.meta.json`
```

### Similarly it can be downloaded as an xls
<!-- Important that eval = FALSE here, otherwise building the vignette downloads data from the ONS API, which is not desirable. -->
```{r , eval = FALSE, include = TRUE}
ons_datasets_setup(monstr_pipeline_defaults()) %>%
	ons_dataset_by_id("weekly-deaths-local-authority") %>%
	ons_download(format="xls")
```


### Specific versions can be selected.
<!-- Important that eval = FALSE here, otherwise building the vignette downloads data from the ONS API, which is not desirable. -->
```{r , eval = FALSE, include = TRUE}
datasets <- ons_datasets_setup(monstr_pipeline_defaults())
## get the metadata about v4 of the time-series edition of weekly-deaths-local-authority dataset.
wdla4_meta <- datasets %>% ons_dataset_by_id("weekly-deaths-local-authority", edition="time-series", version=4)

# download it
wdla4_meta  %>% 
  monstr_pipeline_defaults() %>% 
  ons_download(format="csv")

# Or get the latest
wdla_latest <- datasets %>% ons_dataset_by_id("weekly-deaths-local-authority", edition="time-series")


# csv for the web meta data about the schema of the data.
wdla_latest %>% ons_download(format="csv")

```
