---
title: "Using the monstR package to estimate the mortality rate in England and Wales"
author: "Sebastien Peytrignet"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mortality rate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(monstR)
library(tidyverse)
library(data.table)
library(plotly)
library(scales)
library(ggplot2)
```

## Vignette background

The `monstR` package interacts with the UK's Office for National Statistics (ONS) API, enabling us to easily extract their datasets for analysis. In this vignette, we will demonstrate how to use `monstR` to download the following ONS tables:

- Mid-year population estimates by region, age and gender
- Deaths registered weekly in England and Wales by region

We will then merge both of those datasets to compute a time-series of mortality rates in England and Wales by sub-region. We will also plot those statistics.

## Find identifiers for desired ONS datasets

The function `ons_available_datasets()` is useful to find out what datasets are available through the ONS API.

```{r}
datasets <- ons_available_datasets()
```


Each dataset is associated with an identifier, which we can retrieve in the following way. First, we will extract the identifier for the desired population dataset ('Mid-year population estimates by region, age and gender').

```{r}
population_id <- datasets %>% 
  filter(str_detect(tolower(title),'population estimates for uk')) %>% 
  pull(id)

population_id

datasets %>%
  filter(id %in% population_id) %>%
  select(title)
```

Then, we will do the same for our desired dataset recording deaths by region.

```{r}
mortality_id <- datasets %>% 
  filter(str_detect(tolower(title),'deaths')) %>%
  filter(str_detect(tolower(title),'by region')) %>%
  pull(id)

mortality_id

datasets %>%
  filter(id %in% mortality_id) %>%
  select(title)
```

## Browse editions and versions available for each dataset

ONS datasets are usually associated with multiple editions and versions. Different editions may contain different variables or be presented in different formats, while versions usually refresh or update the content with new data points. Before downloading a dataset with `monstR`, it is recommended to know which edition and version you are interested in. 

The function `ons_available_editions` returns available editions, while `ons_available_versions` return available versions.

We will check the editions and versions associated with the population dataset.

```{r}
### Editions available for the population dataset
ids_and_editions_pop <- map(population_id, ons_available_editions) %>% 
  set_names(population_id) %>% 
  bind_rows(.id='id') %>%
  mutate(.,id_edition=paste(id,edition,sep="-"))

ids_and_editions_pop

### Versions available for each edition
ids_and_editions_and_versions_pop <- mapply(id=ids_and_editions_pop$id,
                                        edition=ids_and_editions_pop$edition,
                                        ons_available_versions)
names(ids_and_editions_and_versions_pop) <- ids_and_editions_pop$id_edition

ids_and_editions_and_versions_pop
```

From the dataset with identifier `mid-year-pop` we would like to download edition `mid-2018-april-2019-geography` and version `1`.

We will do the same for the deaths dataset.

```{r}
### Editions available for the deaths dataset
ids_and_editions_deaths <- map(mortality_id, ons_available_editions) %>% 
  set_names(mortality_id) %>% 
  bind_rows(.id='id') %>%
  mutate(.,id_edition=paste(id,edition,sep="-"))

ids_and_editions_deaths

### Versions available for each edition
ids_and_editions_and_versions_deaths <- mapply(id=ids_and_editions_deaths$id,
                                        edition=ids_and_editions_deaths$edition,
                                        ons_available_versions)
names(ids_and_editions_and_versions_deaths) <- ids_and_editions_deaths$id_edition

ids_and_editions_and_versions_deaths
```

From the dataset with identifier `weekly-deaths-local-authority` we would like to download edition `2010-19` and version `1`.

## Download both ONS datasets

After inspection in the previous step, we have decided which edition and version pair we would like to download for each dataset.

The following command sets up the pipeline using the `monstR` default settings.

```{r}
set_up_df <- monstr_pipeline_defaults() %>% 
  ons_datasets_setup()
```

We indicate the edition and version to function `ons_dataset_by_id`, while 
`ons_download` downloads the data. Additional functions `monstr_read_file`, `monstr_clean` and `monstr_write_clean` read, clean and write the desired datasets.

We can now download our population dataset.

```{r echo=T, results='hide'}
set_up_df %>%
ons_dataset_by_id(id=population_id,edition="mid-2019-april-2020-geography",version=1)  %>%
  ons_download(format="csv") %>%
  monstr_read_file() %>%  
  monstr_clean() %>%
  monstr_write_clean(format="all")
```

And our deaths dataset.

```{r echo=T, results='hide'}
set_up_df %>%
  ons_dataset_by_id(id=mortality_id,edition="2010-19",version=1)  %>%
  ons_download(format="csv") %>%
  monstr_read_file() %>%  
  monstr_clean() %>%
  monstr_write_clean(format="all")
```

## Import and clean population dataset

The cleaned data can be found in your folder under `data/clean/ons/`.

```{r}
setwd('..')

ons_midyear_pop <- fread("data/clean/ons/mid-year-pop-est/mid-2019-april-2020-geography/mid-year-pop-est-v1.csv", header=TRUE, sep=",", check.names=T) %>%
  rename(.,population=v4_0)
```

For this analysis, we will keep the number of deaths for all ages and both genders in each year and region.

```{r}
ons_midyear_pop_total <- filter(ons_midyear_pop,age=="Total"&sex=="All") %>%
  select(.,population,calendar_years,admin_geography) %>%
  arrange(.,admin_geography,desc(calendar_years))

knitr::kable(head(ons_midyear_pop_total, 10))
```

## Import and clean deaths dataset

```{r}
setwd('..')

ons_weekly_deaths_region <- fread("data/clean/ons/weekly-deaths-region/2010-19/weekly-deaths-region-v1.csv", header=TRUE, sep=",", check.names=T) %>%
  rename(.,nr_deaths=v4_1) %>%
  arrange(.,admin_geography,desc(calendar_years))

knitr::kable(head(ons_weekly_deaths_region, 10))
```

This data is currently presented as a weekly time series, with one row per week. We will aggregate it to turn into a yearly time series, thus matching the ONS population dataset.

```{r}
ons_weekly_deaths_region <- as.data.table(ons_weekly_deaths_region)

ons_weekly_deaths_region_annual <- ons_weekly_deaths_region[, list(
  geography  = first(geography),
  nr_deaths=sum(nr_deaths,na.rm=TRUE)),
                         by = list(calendar_years,admin_geography)]

knitr::kable(head(ons_weekly_deaths_region_annual, 10))
```

## Merge datasets

We are now ready to merge our population data into our deaths data, which will allow us to compute a new variable: the death rate per 100,000 residents for a given region and year.

```{r}
ons_weekly_deaths_region_annual <- left_join(ons_weekly_deaths_region_annual,
                                             ons_midyear_pop_total,
                                             by=c("calendar_years" = "calendar_years",
                                                  "admin_geography" = "admin_geography"))

ons_weekly_deaths_region_annual <- mutate(ons_weekly_deaths_region_annual,
                                          deaths_per_100K=nr_deaths/population*100000)

knitr::kable(head(ons_weekly_deaths_region_annual, 10))
```

## Present results in a chart

Having used the `monstR` package to download our required ONS tables and processed the data, we are ready to display the mortality rate by year according to region.

```{r show_figure, fig.height = 4, fig.width = 7, dpi=120}
ons_weekly_deaths_region_annual$bold <- ifelse(ons_weekly_deaths_region_annual$geography=="England and Wales",1,0)

ggplot(ons_weekly_deaths_region_annual,
                    aes(x=factor(calendar_years), y=deaths_per_100K, group=geography)) +
  geom_line(aes(color=geography,size = factor(bold)))+
  geom_point(aes(color=geography)) +
  theme(text = element_text(size = 10),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.key=element_blank()) +
  ggtitle("Mortality rate by region in England and Wales") +
  xlab("Year") + ylab("Number of deaths per 100,000") +
  scale_color_brewer(palette="Set3",name = "Region") +
  scale_size_manual(values = c(0.5,1.25),guide = "none") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",",
                                                 scientific = FALSE))
```